---
title: "Class 12"
author: "Natalie Ogg A91030809"
format: pdf
---

## Section 4: Population Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

## Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

How many samples do we have? ˆ


```{r}
expr <- read.table("expression data.txt")
head(expr)
```


## Sample sizes: 
```{r}
nrow(expr)
table(expr$geno)
```





Let's make a box plot!

```{r}
library(ggplot2)
ggplot(expr, aes(geno, exp, fill=geno)) + geom_boxplot(notch=TRUE) 
```


```{r}

box <- boxplot(exp ~ geno, data = expr, notch = TRUE)
box
```



```{r}
median(box$stats[,1])
median(box$stats[,2])
median(box$stats[,3])
```


## Median expression levels:

A/A: 31.2

A/G: 25.1

G/G: 20.1


## Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

Based on the boxplot and median values, it appears that the A allele has a positive correlation with ORMDL3 gene expression. A/A genotypes result in higher gene expression than G/G genotypes, with heterozygosity resulting in medium levels of expression. 



