---
title: "Class09_halloween"
author: "Natalie Ogg A91030809"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

## Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
## Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

Using `as.logical` converts 1 and 0 to T and F. 

```{r}
as.logical(candy$chocolate)
candy[as.logical(candy$chocolate),]
```


## Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite is Reese's Peanut Butter cup. 

```{r}
candy["Reese's Peanut Butter cup",]$winpercent
```


## Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

## Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library("skimr")
skim(candy)
```




## Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Win percent is on a different scale. 


## Q7. What do you think a zero and one represent for the candy$chocolate column?

Whether they are chocolatey or not. 


## Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```



```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent) + geom_histogram(binwidth=10)
```

## Q9. Is the distribution of winpercent values symmetrical?

No, the peak is to the left. 


## Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
Slightly above 50%. 


## Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choco <- as.logical(candy$chocolate)
candy[choco,"winpercent"]
```

```{r}
fruity <- as.logical(candy$fruit)
candy[fruity, "winpercent"]
```

```{r}
choc.win <- mean(candy[choco, "winpercent"])
fruity.win <- mean(candy[fruity, "winpercent"])
choc.win
fruity.win
```

Chocolate is more popular than fruity on average. 


## Q12. Is this difference statistically significant?

```{r}
t.test(candy[choco, "winpercent"], candy[fruity, "winpercent"])
```
The p-value is way below statistically significant. 


## Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```

Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

## Q15. Make a first barplot of candy ranking based on winpercent values.

## Q16.
```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```


## Q17. What is the worst ranked chocolate candy?

Sixlets

## Q18. What is the best ranked fruity candy?

Starbursts



```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```


## Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Reese's Miniatures (clearest when looking at the graph, but this code helps get numbers on it)


## Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

## Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are the most negatively correlated. Bar and pluribus are also quite negative. 


## Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are most positively correlated. Chocolate and pricepercent are strongly correlated, and chocolate and bar are also common. 



# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```


```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```


```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


```{r}
library(plotly)

```

```{r}
#ggplotly(p)
```



```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


## Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus. This makes sense because fruity, hard, and pluribus tend to be correlated, meaning many candies share those multiple characteristics as we saw using corrplot. 


